---
title: "Pseudotemporal ordering of cells from the distal lung epithelium"
author: "Kieran Campbell"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_align: center
    toc: yes
  pdf_document:
    toc: yes
---

In this vignette we apply `embeddr` to the [Quake et al.](http://www.nature.com/nature/journal/v509/n7500/full/nature13173.html) dataset. For copyright reasons we do not include the raw data in the package. It can be downloaded from [here](http://www.nature.com/nature/journal/v509/n7500/extref/nature13173-s4.txt) and converted to an `SCESet` using [the accompanying vignette](https://github.com/kieranrcampbell/embeddr/blob/master/vignettes/quake_to_scater.Rmd). 

In this example we have saved it to `~/oxford/embeddr/data/lsce.Rdata`, though this should be changed to where you save it.

```{r load-lib, message=FALSE, warning=FALSE}
library(plyr)
library(dplyr)
library(devtools)
library(data.table)
library(ggplot2)
library(reshape2)
library(splines)
library(corrplot)
library(scater)
library(gplots)
library(NMF)
library(monocle)
library(embeddr)

set.seed(123)
```

```{r load-data, cache=TRUE}
## load object 'lsce' 
data_path <- '~/oxford/embeddr/data/lsce.Rdata'
#data_path <- '/net/isi-scratch/kieran/embeddr/data/lsce.Rdata'
load(data_path)
lsce@lowerDetectionLimit <- 0 # from paper
```

## Building the embedding


Next we want to fit a noise model to find highly variable genes:
```{r noise-model, cache=TRUE, fig.align='center'}
Lmeans <- rowMeans(exprs(lsce))
Lvars <- apply(exprs(lsce), 1, var)

to_use <- Lmeans > 0.5 & Lvars > 0 

LCV <- Lvars[to_use] / Lmeans[to_use]^2

df_var <- data.frame(CV2 = LCV, m=Lmeans[to_use])

fit_loglin <- nls(CV2 ~ a + b / m, data = df_var, start=c(a=80, b=5)) 

f <- function(x) coef(fit_loglin)[1] + coef(fit_loglin)[2] / x

is_het <- LCV > 1.5 * predict(fit_loglin)
df_var$is_het <- is_het

ggplot(df_var, aes(x=m, y=CV2, color=is_het)) + geom_point() +
  stat_function(fun=f, color='black') + theme_bw()

het_genes <- names(which(is_het))
het_indices <- match(het_genes, rownames(lsce))
```

Now we can apply laplacian eigenmaps using a nearest-neighbour graph:

```{r le1, cache=TRUE, fig.align='center', message=FALSE, warning=FALSE}
lsce <- embeddr(lsce, genes_for_embedding = het_indices, nn = 10)

phenoData(lsce)$cluster <- pData(lsce)$putative_cell_type
plot_embedding(lsce, color_by = 'putative_cell_type')
plot_graph(lsce)
```


```{r plot-markers-in-space, cache=TRUE, message= FALSE, warning=FALSE}
plot_embedding(lsce, color_by = 'putative_cell_type', 
               plot_genes = c('Sox9','Cited2','Hes1','Fzd2'))
```

It is obvious we have two very separate trajectories; let's consider the pseudotemporal ordering of one only:

```{r cluster-plot, cache=TRUE, fig.align='center', message=FALSE}
set.seed(123)
lsce <- cluster_embedding(lsce, method='mm')#, k = 4)
plot_embedding(lsce)

cl1 <- pData(lsce)$cluster %in% 1:3
lsce_13 <- lsce[,cl1]

lsce_13 <- fit_pseudotime(lsce_13)
lsce_13_tmp <- lsce_13
phenoData(lsce_13_tmp)$cluster <- plyr::mapvalues(pData(lsce_13_tmp)$cluster,
                                                 from=1:3,
                                                 to=c('AT1','BP','AT2'))

plot_embedding(lsce_13_tmp)
plot_pseudotime_density(lsce_13_tmp, color_by = 'cluster')
```

We can also plot genes in pseudotime:

```{r plot-ps, cache=TRUE, fig.align='center'}
plot_in_pseudotime( lsce_13['Fabp5',] )
```

It appears cluster 2 corresponds to `BP` cells, while cluster 1 corresponds to `AT1` and cluster 3 corresponds to `AT2` cells. We'd ideally like to have a pseudotemporal ordering ofr cells as they transition from `BP` to `AT1` and a second transitioning from `BP` to `AT2`. Let's start with the first. From visual inspection there also appear to be some outlier cells mis-classified. Let's remove them before continuing also.

```{r ps-1, cache=TRUE, fig.align='center'}
lsce_bp_at1 <- lsce[, pData(lsce)$cluster %in% c(1,2) ]
lsce_bp_at1 <- fit_pseudotime(lsce_bp_at1)
lsce_bp_at1 <- reverse_pseudotime(lsce_bp_at1)
plot_embedding(lsce_bp_at1)
```

## Finding genes differentially expressed across pseudotime

### BP to AT1 transition

Now we can plot genes just for this in pseudotime:
```{r remove-zero-var, cache=TRUE}
gene <- 'Fabp5'
plot_in_pseudotime( lsce_bp_at1[gene,])

## Remove genes expressed in less than 20% of cells
n_cells_expressed <- rowSums(is_exprs(lsce_bp_at1))
keep_gene <- n_cells_expressed > 0.2 * dim(lsce_bp_at1)[2]
lsce_bp_at1 <- lsce_bp_at1[keep_gene,]
```

We can also fit a pseudotime cubic smoothing spline regression to `Fabp5`:

```{r plot-fap, cache=TRUE, fig.align='center', fig.width=5.5, fig.height=3.5, warning=FALSE, message=FALSE}
model <- fit_pseudotime_model(lsce_bp_at1, gene)
null_model <- fit_null_model(lsce_bp_at1,  gene)
p_val <- compare_models(model, null_model)
plot_pseudotime_model(lsce_bp_at1[gene,], model) + ggtitle('Fabp5 expression in pseudotime')
plot_pseudotime_model(lsce_bp_at1[gene,], null_model) + ggtitle('Fabp5 expression under null model')
print(paste('P-value: ', p_val))
```

We can then perform the same tests across all genes:

```{r vary-sig-at1, cache=TRUE}
lsce_bp_at1@lowerDetectionLimit <- 0
diff_gene_test <- pseudotime_test(lsce_bp_at1)
```

Plot the p-values:

```{r plot-pval, fig.align='center', cache=TRUE, fig.width=5, fig.height=3.5, warning=FALSE}
## correct differential gene test
qplot(diff_gene_test$p_val, binwidth = 0.01) + theme_bw() + xlab('p-value') 
qplot(diff_gene_test$q_val, binwidth = 0.01) + theme_bw() + xlab('corrected p-value')
alpha <- 0.01
sig_genes <- as.character(diff_gene_test$gene[diff_gene_test$q_val < alpha])
lsce_sig <- lsce_bp_at1[sig_genes,]
```

```{r print-at1, fig.align='center', fig.height=6, fig.width=9, cache=TRUE}
lsce_sig <- reverse_pseudotime(lsce_sig)
plot_pseudotime_model(lsce_sig, facet_wrap_scale = 'free_y')
```

Plot correlations of genes across pseudotime:
```{r plot-corr, fig.align='center', fig.width=9, fig.height=9, cache=TRUE}
pe <- predicted_expression(lsce_sig)
```

Look at directionality of genes by hierarchical clustering:
```{r corr-clust, fig.align='center', cache=TRUE, fig.width=9}
pcor <- cor(pe)
dist_cor <- 1 - pcor / 2 
hc <- hclust(as.dist(dist_cor))
plot(hc)
```

Finally plot genes going up or down in pseudotime:
```{r gene-up, message=FALSE, fig.align='center', cache=TRUE, fig.width=10, fig.height=4}
n_cuts <- 3
gene_classes <- cutree(hc, n_cuts)
      
df_gene <- data.frame(gene=colnames(pe), class=gene_classes)

pe <- data.frame(scale(pe)) ## scaled pst-by-gene
pe$pseudotime <- pseudotime(lsce_sig)
## save(pe, df_gene, file='~/pe.Rdata')

pe_melted <- melt(pe, id.vars='pseudotime', value.name='expression', variable.name='gene')
pe_melted <- inner_join(pe_melted, df_gene)

## we want to plot the mean expression rather than all of it (messy)
gp_groups <- group_by(pe_melted, class, pseudotime)
mean_expr_per_group <- dplyr::summarise(gp_groups, mean_expr = mean(expression))
pe_melted <- inner_join(pe_melted, mean_expr_per_group)
#pe_melted <- arrange(pe_melted, gene, pseudotime)

ggplot(pe_melted) + geom_line(aes(x=pseudotime, y=mean_expr), color='black') +
  facet_wrap(~ class) + geom_point(aes(x=pseudotime, y=expression, color=gene), alpha=0.5) + 
  theme_bw() + ylab('Expression') + guides(color=FALSE)

```

These genes are selectively turned off in AT1 cells, meaning they're markers for AT2 cells. Let's see how what we found compares to those in the original paper:

```{r marker-at2, cache=TRUE}
markers2 <- c('Lcn2', 'Hc','Trf','Lyz2','S100g','Lyz1','Fabp5',
            'Lamp3','Cd36','Scd1','Sftpb','Slc34a2','Abca3','Sftpa1',
            'Egfl6','Soat1','Bex2','Muc1','Sftpc')
pseudotime_markers_at2 <- as.character(sig_genes)

at2_both <- intersect(markers2, pseudotime_markers_at2)
at2_prev <- setdiff(markers2, pseudotime_markers_at2)
at2_embeddr <- setdiff(pseudotime_markers_at2, markers2)
print(c(length(at2_both), length(at2_prev), length(at2_embeddr)))
print(at2_both)
print(at2_prev)
print(at2_embeddr)
```

And plot pseudotime density:

```{r cell-density-1, cache=TRUE, fig.align='center'}
lsce_bp_at1_tmp <- lsce_bp_at1
phenoData(lsce_bp_at1_tmp)$cluster <- plyr::mapvalues(pData(lsce_bp_at1_tmp)$cluster,
                                                 from=1:2,
                                                 to=c('AT1','BP'))

plot_pseudotime_density(lsce_bp_at1_tmp, reverse = TRUE)
```

### BP to AT2 transition

```{r ps-2, cache=TRUE, fig.align='center'}
lsce_bp_at2 <- lsce[, pData(lsce)$cluster %in% c(3,2)]
lsce_bp_at2 <- fit_pseudotime(lsce_bp_at2)
plot_embedding(lsce_bp_at2)
```


```{r remove-zero-var2, cache=TRUE}
gene <- 'Fabp5'
plot_in_pseudotime( lsce_bp_at2[gene,])

## Remove genes expressed in less than 20% of cells
n_cells_expressed <- rowSums(is_exprs(lsce_bp_at2))
keep_gene <- n_cells_expressed > 0.2 * dim(lsce_bp_at2)[2]
lsce_bp_at2 <- lsce_bp_at2[keep_gene,]
print(lsce_bp_at2)
```

We can also fit a pseudotime cubic smoothing spline regression to `Fabp5`:

```{r plot-fap2, cache=TRUE, fig.align='center', fig.width=5.5, fig.height=3.5}
model <- fit_pseudotime_model(lsce_bp_at2, gene)
null_model <- fit_null_model(lsce_bp_at2,  gene)
p_val <- compare_models(model, null_model)
plot_pseudotime_model(lsce_bp_at2[gene,], model) + ggtitle('Fabp5 expression in pseudotime')
print(paste('P-value: ', p_val))
```

We can then perform the same tests across all genes:

```{r vary-sig-at2, cache=TRUE, message=FALSE}
lsce_bp_at2@lowerDetectionLimit <- 0
diff_gene_test <- pseudotime_test(lsce_bp_at2, n_cores = 2)

```

Plot the p-values:

```{r plot-pval2, fig.align='center', cache=TRUE, fig.width=5, fig.height=3.5, message=FALSE}
## correct differential gene test
diff_gene_test <- dplyr::filter(diff_gene_test, p_val < 1)
diff_gene_test$q_val <- p.adjust(diff_gene_test$p_val, method='BH')
qplot(diff_gene_test$p_val, binwidth = 0.01) + theme_bw() + xlab('p-value') 
qplot(diff_gene_test$q_val, binwidth = 0.01) + theme_bw() + xlab('corrected p-value')
alpha <- 0.01
sig_genes <- diff_gene_test$gene[diff_gene_test$q_val < alpha]
lsce_sig <- lsce_bp_at2[sig_genes,]

```

```{r plot-at2, fig.align='center', fig.height=9, fig.width=9, cache=TRUE}
plot_pseudotime_model(lsce_sig, facet_wrap_scale = "free_y")
```

Plot correlations of genes across pseudotime:
```{r plot-corr2, fig.align='center', fig.width=9, fig.height=9, message=FALSE}
pe <- predicted_expression(lsce_sig)
```

Look at directionality of genes by hierarchical clustering:
```{r corr-clust2, fig.align='center', cache=TRUE, fig.width=9}
pcor <- cor(pe)
dist_cor <- 1 - pcor / 2 
hc <- hclust(as.dist(dist_cor))
plot(hc)
```


Finally plot genes going up or down in pseudotime:
```{r gene-up2, fig.align='center', cache=TRUE, fig.width=9, fig.height=4, message=FALSE, warning=FALSE}
n_cuts <- 3
gene_classes <- cutree(hc, n_cuts)

lines <- c()
for(i in 1:n_cuts) {
  genes_i <- names(which(gene_classes == i))
  lines <- c(lines, (c(paste('class',i), '\n')))
  lines <- c(lines, genes_i)
}

# fileConn <- file('~/oxford/embeddr/gene_classes.txt')
# writeLines(lines, fileConn)
# close(fileConn)

df_gene <- data.frame(gene=colnames(pe), class=gene_classes)

pe <- data.frame(scale(pe)) ## scaled pst-by-gene
pe$pseudotime <- pseudotime(lsce_sig)
## save(pe, df_gene, file='~/pe.Rdata')

pe_melted <- melt(pe, id.vars='pseudotime', value.name='expression', variable.name='gene')
pe_melted <- inner_join(pe_melted, df_gene)

## we want to plot the mean expression rather than all of it (messy)
gp_groups <- group_by(pe_melted, class, pseudotime)
mean_expr_per_group <- dplyr::summarise(gp_groups, mean_expr = mean(expression))
pe_melted <- inner_join(pe_melted, mean_expr_per_group)
## pe_melted <- arrange(pe_melted, gene, pseudotime)

ggplot(pe_melted) + geom_line(aes(x=pseudotime, y=mean_expr), color='red') +
  facet_wrap(~ class) + stat_density2d(aes(x=pseudotime, y=expression), n=100) +
  theme_bw() + ylab('Expression')

```



```{r at1-markers, cache=TRUE}
markers1 <- c('Aqp5','Pdpn','Rtkn2','Ager','Emp2','Cav1','Clic5','Lmo7',
             'S100a6','Col4a3','Akap5','Cryab','Sdpr','S100a14')
pseudotime_markers_at1 <- names(gene_classes[gene_classes == 2])
pseudotime_markers_at2_upreg <- names(gene_classes[gene_classes == 1])

at1_both <- intersect(markers1, pseudotime_markers_at1)
at1_original <- setdiff(markers1, pseudotime_markers_at1)
at1_embeddr <- setdiff(pseudotime_markers_at1, markers1)

print(c(length(at1_both), length(at1_original), length(at1_embeddr)))

all_pseudotime_at2 <- union(pseudotime_markers_at2_upreg, pseudotime_markers_at2)
all_at2_both <- intersect(markers2, all_pseudotime_at2)
all_at2_original <- setdiff(markers2, all_pseudotime_at2)
all_at2_embeddr <- setdiff(all_pseudotime_at2, markers2)

print(c(length(all_at2_both), length(all_at2_original), length(all_at2_embeddr)))
```

### Bringing it all together

```{r big-heatmap, fig.align='center', fig.width=9, fig.height=4.5}
agree_1 <- intersect(pseudotime_markers_at1, markers1)
agree_2 <- intersect(all_pseudotime_at2, markers2)

paper_only_1 <- setdiff(markers1, pseudotime_markers_at1)
paper_only_2 <- setdiff(markers2, all_pseudotime_at2)

embeddr_only_1 <- setdiff(pseudotime_markers_at1, markers1)
embeddr_only_2 <- setdiff(all_pseudotime_at2, markers2)

all_pseudotime_markers <- c(agree_1, paper_only_1, embeddr_only_1,
                            agree_2, paper_only_2, embeddr_only_2)

previously_identified <- c(rep('Both', length(agree_1)),
                           rep('Prev', length(paper_only_1)),
                           rep('New', length(embeddr_only_1)),
                           rep('Both', length(agree_2)),
                           rep('Prev', length(paper_only_2)),
                           rep('New', length(embeddr_only_2)))
# prev_colors <- plyr::mapvalues(previously_identified, from=c('Prev','New'), to=c('darkgrey','lightgrey'))

cell_type <- pData(lsce_13)$putative_cell_type[order(pseudotime(lsce_13))]
# cell_colors <- plyr::mapvalues(cell_type, from=c('AT1','BP','AT2'), to=c('orange','black','darkred'))

mm <- exprs(lsce_13[all_pseudotime_markers,order(pseudotime(lsce_13))])

# save(mm, cell_type, cell_colors, previously_identified, prev_colors, file='~/for_heatmap.rdata')
# 
# my_palette <- colorRampPalette(c("darkblue","white","red"))(n = 100)
# heatmap.2(mm, Colv=FALSE, Rowv=FALSE, col=my_palette, dendrogram='none', trace='none',
#           density.info='none', lhei=c(1, 3), lwid=c(2, 7), ColSideColors=cell_colors, RowSideColors=prev_colors)
# legend("top",      # location of the legend on the heatmap plot
#     legend = c('AT1','BP','AT2'),
#     col = c('orange','black','darkred'),
#     lty= 1,             # line style
#     lwd = 5            # line width
# )
aheatmap(mm, Rowv = NA, Colv = NA,
         annRow = data.frame('Time discovered' = previously_identified),
         annCol = data.frame('Cell Type' = cell_type))
```


## Compare to monocle

```{r pull-out-sce}
save(lsce_bp_at1, pseudotime_markers_at2, file="~/delete_me.Rdata")
```

```{r compare-monocle, cache=TRUE}
# lcds <- toCellDataSet(lsce)
# lcds <- setOrderingFilter(lcds, het_genes)
# lcds2 <- lcds # logs data for you
# exprs(lcds2) <- 2^exprs(lcds) - 1
# 
# lcds2 <- reduceDimension(lcds2, use_irlba = FALSE)
# lcds2 <- orderCells(lcds2, num_paths = 2)
# plot_spanning_tree(lcds2, color_by = 'putative_cell_type')
# 
# ## BP - AT1 transition
# cds_at_bp1 <- lcds2[,lcds$putative_cell_type %in% c('BP','AT1')]
# #cds_at_bp1 <- reduceDimension(cds_at_bp1, use_irlba=FALSE)
# cds_at_bp1 <- orderCells(cds_at_bp1, num_paths = 1)
# plot_spanning_tree(cds_at_bp1, color_by = 'putative_cell_type')
# 
# ## BP - AT2 transition
# cds_at_bp2 <- lcds2[,lcds$putative_cell_type %in% c('BP','AT2')]
# #cds_at_bp1 <- reduceDimension(cds_at_bp1, use_irlba=FALSE)
# cds_at_bp2 <- orderCells(cds_at_bp2, num_paths = 1)
# plot_spanning_tree(cds_at_bp2, color_by = 'putative_cell_type')
# 
# n_cells_expressed <- rowSums(exprs(cds_at_bp1) > 0)
# keep_gene <- n_cells_expressed > 0.2 * dim(cds_at_bp1)[2]
# cds <- cds_at_bp1[keep_gene,]
# 
# exprs(cds) <- exprs(cds) + 1
# cds@lowerDetectionLimit <- log10(1)
# # exprs(cds) <- exprs(cds) + 1
# diff_test_res <- differentialGeneTest(cds)
# 
# n_cells_expressed <- rowSums(exprs(cds_at_bp2) > 0)
# keep_gene <- n_cells_expressed > 0.2 * dim(cds_at_bp2)[2]
# cds <- cds_at_bp2[keep_gene,]
# exprs(cds) <- exprs(cds) + 1
# diff_test_res <- differentialGeneTest(cds)

```



