---
title: "Pseudotemporal ordering of cells from the distal lung epithelium"
author: "Kieran Campbell"
date: "03/24/2015"
output: html_document
---

Initial data processing: first we read in the data and remove bulk transcripts:

```{r load-lib, message=FALSE}
library(dplyr)
library(devtools)
library(data.table)
library(ggplot2)
library(splines)
library(VGAM)
library(corrplot)
library(scater)
embeddr_path <-  "/net/isi-scratch/kieran/embeddr/embeddr" # '~/oxford/embeddr/embeddr' #
load_all(embeddr_path)

set.seed(123)
```

```{r load-data, cache=TRUE}
## load object 'sce' 
load('data/lung_sce.RData.gz')
```

Next we want to fit a noise model to find highly variable genes:
```{r noise-model, cache=TRUE, fig.align='center'}
Lmeans <- rowMeans(exprs(sce))
Lvars <- apply(exprs(sce), 1, var)

to_use <- Lmeans > 0.5 & Lvars > 0 

LCV <- Lvars[to_use] / Lmeans[to_use]^2

df_var <- data.frame(CV2 = LCV, m=Lmeans[to_use])

fit_loglin <- nls(CV2 ~ a + b / m, data = df_var, start=c(a=80, b=5)) 

f <- function(x) coef(fit_loglin)[1] + coef(fit_loglin)[2] / x

is_het <- LCV > 1.5 * predict(fit_loglin)
df_var$is_het <- is_het

ggplot(df_var, aes(x=m, y=CV2, color=is_het)) + geom_point() +
  stat_function(fun=f, color='black') + theme_bw()

het_genes <- names(which(is_het))
het_indices <- match(het_genes, rownames(sce))

sce_var <- sce[het_indices,]
```

Now we can apply laplacian eigenmaps using a nearest-neighbour graph:

```{r le1, cache=TRUE, fig.align='center', message=FALSE, warning=FALSE}
W <- weighted_graph(sce_var, metric='correlation', nn=10)
M <- laplacian_eigenmap(sce, W, type='unorm')

phenoData(sce)$cluster <- pData(sce)$putative_cell_type
plot_embedding(sce)
plot_graph(sce, W)
```

It is obvious we have two very separate trajectories; let's consider the pseudotemporal ordering of one only:

```{r cluster-plot, cache=TRUE, fig.align='center', message=FALSE}
sce <- cluster_embedding(sce, method='mm')
plot_embedding(sce)

cl1 <- pData(sce)$cluster %in% 1:3
sce_13 <- sce[,cl1]

sce_13 <- fit_pseudotime(sce_13)

plot_embedding(M1)
```

We can also plot genes in pseudotime:

```{r plot-ps, cache=TRUE, fig.align='center'}
plot_in_pseudotime( sce_13 , genes='Fabp5' )
```

It appears cluster 2 corresponds to `BP` cells, while cluster 1 corresponds to `AT1` and cluster 3 corresponds to `AT2` cells. We'd ideally like to have a pseudotemporal ordering ofr cells as they transition from `BP` to `AT1` and a second transitioning from `BP` to `AT2`. Let's start with the first. From visual inspection there also appear to be some outlier cells mis-classified. Let's remove them before continuing also.

```{r ps-1, cache=TRUE, fig.align='center'}
M_bp_at1 <- filter(M, cluster %in% 1:2, component_1 < -0.045)
M_bp_at1 <- fit_pseudotime(M_bp_at1)
plot_embedding(M_bp_at1)
```

Now we need to get the gene expression matrix confined to the cells we've identified:
```{r remove-zero-var, cache=TRUE}
bp_at1 <- M$cluster %in% 1:2 & M$component_1 < -0.045
dd_bp_at1 <- data.frame(d_het[, bp_at1])
plot_in_pseudotime( M_bp_at1, dd_bp_at1, genes='Fabp5' )

## Remove zero variance genes
print(dim(dd_bp_at1))
toremove <- which(apply(dd_bp_at1, 1, function(x) var(x) == 0))
dd_bp_at1 <- dd_bp_at1[-toremove,]
dd_bp_at1 <- t(dd_bp_at1)
rownames(dd_bp_at1) <- NULL
print(dim(dd_bp_at1))
```

We can also fit a pseudotime cubic smoothing spline regression to `Fabp5`:

```{r plot-fap, cache=TRUE, fig.align='center', fig.width=5.5, fig.height=3.5}
t <- M_bp_at1$pseudotime
y <- dd_bp_at1[,'Fabp5']
min_expr <- 0 # see paper

model <- fit_pseudotime_model(y, t, min_expr)
null_model <- fit_null_model(y, min_expr)
p_val <- compare_models(model, null_model)
plot_pseudotime_model(model, y, t, min_expr) + ggtitle('Fabp5 expression in pseudotime')
print(p_val)
```

We can then perform the same tests across all genes:

```{r vary-sig-at1, cache=TRUE}
test_gene <- function(y, t, min_expr) {
  tryCatch({
  model <- embeddr::fit_pseudotime_model(y, t, min_expr)
  null_model <- embeddr::fit_null_model(y, min_expr)
  return( embeddr::compare_models(model, null_model) )
  }, error = function(e) {
    return( 1 ) # if there's an error, just return 1
  })
}

df_all_genes <- data.frame(d_numeric[bp_at1, ])
toremove <- which(apply(df_all_genes, 2, function(x) var(x) == 0))
df_all_genes_nonzero  <- df_all_genes[,-toremove]
sig_tests <- apply(df_all_genes_nonzero, 2, test_gene, t, min_expr)
```

Plot the p-values:

```{r plot-pval, fig.align='center', cache=TRUE, fig.width=5, fig.height=3.5}
qplot(sig_tests) + theme_minimal()
```

Adjust the p-values to find significantly varying ones:

```{r sig-vary, cache=TRUE}
adj_sigs <- p.adjust(sig_tests, method='BH')

alpha <- 0.005
sig_genes <- names(which(adj_sigs < alpha))
sg <- adj_sigs[adj_sigs < alpha]
```

Plot correlations of genes across pseudotime:
```{r plot-corr, fig.align='center', fig.width=9, fig.height=9}
df_cor <- df_all_genes[,sig_genes]
gene_cor <- cor(df_cor)
corrplot(gene_cor,order='hclust')

```

Look at directionality of genes by hierarchical clustering:
```{r corr-clust, fig.align='center', cache=TRUE, fig.width=9}
dis <- 1 - gene_cor / 2
ddis <- as.dist(dis)

hc <- hclust(ddis)
plot(hc)
```

Finally plot genes going up or down in pseudotime:
```{r gene-up, fig.align='center', cache=TRUE, fig.width=10, fig.height=8}
gene_groups <- cutree(hc, 6)
plot_in_pseudotime(M_bp_at1, t(df_all_genes), sig_genes[gene_groups == 1])
plot_in_pseudotime(M_bp_at1, t(df_all_genes), sig_genes[gene_groups == 2])
plot_in_pseudotime(M_bp_at1, t(df_all_genes), sig_genes[gene_groups == 3])
plot_in_pseudotime(M_bp_at1, t(df_all_genes), sig_genes[gene_groups == 4])
plot_in_pseudotime(M_bp_at1, t(df_all_genes), sig_genes[gene_groups == 5])
plot_in_pseudotime(M_bp_at1, t(df_all_genes), sig_genes[gene_groups == 6])

```

Now let's do the same for the transition into AT2:

```{r ps-2, cache=TRUE, fig.align='center'}
M_bp_at2 <- filter(M, cluster %in% 2:3, component_1 < 0.045)
M_bp_at2 <- fit_pseudotime(M_bp_at2)
plot_embedding(M_bp_at2)
M_bp_at2 <- reverse_pseudotime(M_bp_at2)
```

Now we need to get the gene expression matrix confined to the cells we've identified:
```{r remove-zero-var-2, cache=TRUE}
bp_at2 <- M$cluster %in% 2:3 & M$component_1 < 0.045
dd_bp_at2 <- data.frame(d_het[, bp_at2])
plot_in_pseudotime( M_bp_at2, dd_bp_at2, genes='Fabp5' )

t <- M_bp_at2$pseudotime

## Remove zero variance genes
df_all_genes <- data.frame(d_numeric[bp_at2, ])
toremove <- which(apply(df_all_genes, 2, function(x) var(x) == 0))
df_all_genes_nonzero  <- df_all_genes[,-toremove]
sig_tests <- apply(df_all_genes_nonzero + 1, 2, test_gene, t, min_expr + 1)
```

And another look at the p-values:

```{r plot-pval-2, fig.align='center', cache=TRUE, fig.width=5, fig.height=3.5, warning=FALSE}
qplot(sig_tests) + theme_minimal()
```


```{r sig-vary-2, cache=TRUE}
adj_sigs <- p.adjust(sig_tests, method='BH')
alpha <- 0.005
sig_genes <- names(which(adj_sigs < alpha))
sg <- adj_sigs[adj_sigs < alpha]

```

Plot correlations of genes across pseudotime:
```{r plot-corr-2, fig.align='center', fig.width=9, fig.height=9}
df_cor <- df_all_genes[,sig_genes]
gene_cor <- cor(df_cor)
corrplot(gene_cor,order='hclust')

```

Look at directionality of genes by hierarchical clustering:
```{r corr-clust-2, fig.align='center', cache=TRUE, fig.width=9}
dis <- 1 - gene_cor / 2
ddis <- as.dist(dis)
hc <- hclust(ddis)
plot(hc)

```

```{r gene-pattern-2, fig.align='center', cache=TRUE, fig.width=10, fig.height=8}
gene_groups <- cutree(hc, 3)
plot_in_pseudotime(M_bp_at2, t(df_all_genes), sig_genes[gene_groups == 1])
plot_in_pseudotime(M_bp_at2, t(df_all_genes), sig_genes[gene_groups == 2])
plot_in_pseudotime(M_bp_at2, t(df_all_genes), sig_genes[gene_groups == 3])

```

So we see there are three main classes of genes: early downregulation, late downregulation and upregulation.

